High-dimensional Asymptotics of Transfer Learning for Linear Regression

### 2021-06-12 revision_journal/backup/Jan 29 2021/manuscript.tex
revision wip

\begin{frontmatter}
\title{Hard Parameter Sharing in High-dimensional Linear Regression}
%\title{A sample article title with some additional note\thanksref{t1}}
\runtitle{Hard Parameter Sharing in High-dimensional Linear Regression}
%\thankstext{T1}{A sample additional note to the title.}

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%Only one address is permitted per author. %%
%%Only division, organization and e-mail is %%
%%included in the address.                  %%
%%Additional information can be included in %%
%%the Acknowledgments section if necessary. %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\fnms{Fan} \snm{Yang}\ead[label=e1]{fyang75@wharton.upenn.edu} \thanks{Fan Yang and Hongyang R. Zhang contributed equally.}}
\author[B]{\fnms{Hongyang R.} \snm{Zhang}\ead[label=e2,mark]{hrzhang@northeastern.edu}\samethanks} 
\author[C]{\fnms{Sen} \snm{Wu}\ead[label=e3,mark]{senwu@stanford.edu}} 
\author[A]{\fnms{Weijie J.} \snm{Su}\ead[label=e4,mark]{suw@wharton.upenn.edu} }
\and
\author[C]{\fnms{Christopher} \snm{R\'e}\ead[label=e5,mark]{chrismre@cs.stanford.edu}} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Department of Statistics,
University of Pennsylvania,
\printead{e1,e4}}
\address[B]{Khoury College of Computer Sciences, Northeastern University,
\printead{e2}}
\address[C]{Department of Computer Science, Stanford University,
\printead{e3,e5}}

@@

\begin{abstract}
	Hard parameter sharing for multi-task learning is widely used in empirical research despite the fact that its generalization properties have not been well established in many cases. This paper studies its generalization properties in a fundamental setting: How does hard parameter sharing work given multiple linear regression tasks? We develop new techniques and establish a number of new results in the high-dimensional setting, where the sample size and feature dimension increase at a fixed ratio. First, we show a sharp bias-variance decomposition of hard parameter sharing, given multiple tasks with the same features. Second, we characterize the asymptotic bias-variance limit for two tasks, even when they have arbitrarily different sample size ratios and covariate shifts. We also demonstrate that these limiting estimates for the empirical loss are incredibly accurate in moderate dimensions. Finally, we explain an intriguing phenomenon where increasing one task's sample size helps another task initially by reducing variance but hurts eventually due to increasing bias. This suggests progressively adding data for optimizing hard parameter sharing, and we validate its efficiency in text classification tasks.
\end{abstract}

@@

Hard parameter sharing (HPS) for multi-task learning is widely used in empirical research and goes back to the seminal work of \cite{C97}.
Recent work has revived interests in this approach because it improves performance and reduces the cost of collecting labeled data \cite{R17}.
It is generally applied by sharing the feature layers between all tasks while keeping an output layer for every task.
Often, hard parameter sharing offers two critical advantages if successfully applied.
First, it reduces model parameters since all tasks use the same feature space.
Second, it reduces the amount of labeled data needed from each task by augmenting the entire training dataset.

@@

\section{Hard parameter sharing estimator}\label{sec_HPS}

Consider the transfer learning setting with two data sets. We denote their sample sizes by $n_1$ and $n_2$. For $i=1,2,$ Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates. We assume that the label vector $Y^{(i)} \in \real^{n_i}$ for $X^{(i)}$ follows a linear model plus random noise:
$$Y^{(i)}= X^{(i)}\beta^{(i)} + \varepsilon^{(i)},\quad i=1,2.$$
We study the standard hard parameter sharing architecture: a shared feature representation layer $B\in\real^{p}$ for all datasets and a separate output layer $A_i \in \real$ for every dataset $i$. Then we study the following minimization problem:
\begin{align}\label{eq_tsl}
			f(A, B) = \norm{X^{(1)} B A_1 - Y^{(1)}}^2 + \norm{X^{(2)} B A_2 - Y^{(2)}}^2,
\end{align}
where we abbreviate $A = [A_1, A_2]$. Let $(\hat{A}, \hat{B})$ be the minimizer of $f(A, B)$. We define the hard parameter sharing (HPS) estimator for task $i$ as 
\be\label{def_HPS}\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i,\quad i=1,2.\ee
In this paper, we make the following assumptions on feature covariates $X^{(i)}$ and the noise vectors $\varepsilon^{(i)}$. We remark that all these assumptions are natural in high-dimensional statistics.

@@

@@

%However, to the best of our knowledge, the asymptotic limit of $L(\hat{\beta}_2^{\MTL}(a))$ has never been calculated exactly in any high-dimensional setting.

\section{HPS estimator: Bias-variance tradeoff}\label{sec_same}

\section{HPS estimator: Varying sample sizes} 	\label{sec_sizeratio}

@@

For a fixed $n_2$, we study the intricate bias-variance tradeoff with respect to $n_1$ using the following function of $\rho:=\rho_1+\rho_2$,
$$h(\rho):=\frac{p }{n_2-p}\cdot \frac{\sigma^2}{2d^2}- \frac{\ell(\rho_1p , \rho_2p)}{2d^2}  = \frac{\rho-\rho_2}{(\rho-1)(\rho_2-1)}\cdot \frac{\sigma^2}{2d^2}- \frac{(\rho-\rho_2)^2 (\rho-1)+(\rho-\rho_2)\rho_2}{\rho^2(\rho-1)}  .$$
This function characterizes $L(\hat{\beta}_2^{\STL} )-L(\hat{\beta}_2^{\MTL}(\hat a)) $, which gives the quantitive information transfer from the source task to the target task.
First, we have positive (resp. negative) transfer if and only if $h(\rho)>0$ (resp. $h(\rho)<0$). Moreover, the sign of $h(\rho)$ is determined by the sign of the second order polynomial %which is equivalent to the following inequality:
\be\label{poly1} \left( \frac{1}{\rho_2-1}\cdot \frac{\sigma^2}{2d^2}-1 \right)\rho^2 + ( \rho_2+1)\rho - 2\rho_2 .\ee
Then we observe the following dichotomy.
\begin{enumerate}
\item If $\frac{1}{\rho_2-1}\cdot \frac{\sigma^2}{2d^2}-1 >0$, then the polynomial \eqref{poly1} is positive for all $\rho\in (\rho_2+1,\infty)$, so that the transfer is always positive.

@@

\item If $(1+\rho_2)^2 < 8\rho_2 \left( 1- \frac{1}{\rho_2-1}\cdot \frac{\sigma^2}{2d^2}  \right)$, the polynomial \eqref{poly1} has two positive roots, where one of them is always smaller than $\rho_2+1$. Hence if the larger root, say $\rho_c$, is smaller than $\rho_2+1$, then the transfer is always negative for $\rho\in (\rho_2+1,\infty)$; otherwise, there is a transition from positive transfer to negative transfer as $\rho$ crosses $\rho_c$.
\end{itemize}
\end{enumerate}
Second, we calculate the derivative of $h(\rho)$, and find that its sign  is determined by the sign of the third order polynomial 
\be\label{poly2}
 \frac{1}{\rho_2}\cdot \frac{\sigma^2}{2d^2} \rho^3 - 2(\rho-\rho_2)(\rho-1)(\rho-2)   - (\rho_2-1)\rho .\ee
Then we observe the following dichotomy.
\begin{enumerate}
\item If $\frac{1}{\rho_2}\cdot \frac{\sigma^2}{2d^2} > 2$, the polynomial \eqref{poly2} is always positive for all $\rho\in (\rho_2+1,\infty)$, so that the information transfer always increases as $n_1$ increases.

\item If $\frac{1}{\rho_2}\cdot \frac{\sigma^2}{2d^2} < 2$, the polynomial \eqref{poly2} is positive initially around $\rho=\rho_2+1$, and then becomes negative when $\rho$ crosses its unique root, say $\rho_c$, in $(\rho_2+1,\infty)$. Hence the information transfer achieves the global minimum at $\rho=\rho_c$.
\end{enumerate}

@@

\section{HPS estimator: Covariate shift}\label{sec_diff}

@@

@@

 \section*{Acknowledgements}
Thanks to Edgar Dobriban, Vaggos Chatziafratis, Mayee Chen, and Ruoxuan Xiong for helpful discussions at various stages of this work. Hongyang R. Zhang and Weijie J. Su's contributions were in part supported by the Wharton Dean's Fund for Postdoctoral Research.
We gratefully acknowledge the support of DARPA under Nos. FA86501827865 (SDH) and FA86501827882 (ASED); NIH under No. U54EB020405 (Mobilize), NSF under Nos. CCF1763315 (Beyond Sparsity), CCF1563078 (Volume to Velocity), and 1937301 (RTML); ONR under No. N000141712266 (Unifying Weak Supervision); the Moore Foundation, NXP, Xilinx, LETI-CEA, Intel, IBM, Microsoft, NEC, Toshiba, TSMC, ARM, Hitachi, BASF, Accenture, Ericsson, Qualcomm, Analog Devices, the Okawa Foundation, American Family Insurance, Google Cloud, Swiss Re, the HAI-AWS Cloud Credits for Research program, and members of the Stanford DAWN project: Teradata, Facebook, Google, Ant Financial, NEC, VMWare, and Infosys. The U.S. Government is authorized to reproduce and distribute reprints for Governmental purposes notwithstanding any copyright notation thereon. Any opinions, findings, and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views, policies, or endorsements, either expressed or implied, of DARPA, NIH, ONR, or the U.S. Government.

>>>>

\begin{frontmatter}
%\title{High-dimensional Asymptotics of Linear Regression with Different Covariates and Sample Sizes}
%\title{A sample article title with some additional note\thanksref{t1}}

\title{High-dimensional Asymptotics of Transfer Learning for Linear Regression} 
%with Different Covariates and Sample Sizes}

\runtitle{Linear Regression with Different Covariates and Sample Sizes}
%\thankstext{T1}{A sample additional note to the title.}

\begin{aug}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%Only one address is permitted per author. %%
%%Only division, organization and e-mail is %%
%%included in the address.                  %%
%%Additional information can be included in %%
%%the Acknowledgments section if necessary. %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\author[A]{\fnms{Fan} \snm{Yang}\ead[label=e1]{fyang75@wharton.upenn.edu}}
\author[B]{\fnms{Hongyang R.} \snm{Zhang}\ead[label=e2,mark]{hrzhang@northeastern.edu}}
\author[C]{\fnms{Sen} \snm{Wu}\ead[label=e3,mark]{senwu@stanford.edu}} 
\author[A]{\fnms{Weijie J.} \snm{Su}\ead[label=e4,mark]{suw@wharton.upenn.edu} }
\and
\author[C]{\fnms{Christopher} \snm{R\'e}\ead[label=e5,mark]{chrismre@cs.stanford.edu}} 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% Addresses                                %%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\address[A]{Department of Statistics,
University of Pennsylvania,
\printead{e1,e4}}
\address[B]{Khoury College of Computer Sciences, Northeastern University,
\printead{e2}}
\address[C]{Department of Computer Science, Stanford University,
\printead{e3,e5}}

@@

\begin{abstract}
	{\cor rewrite} Hard parameter sharing for multi-task learning is widely used in empirical research despite the fact that its generalization properties have not been well established in many cases. This paper studies its generalization properties in a fundamental setting: How does hard parameter sharing work given multiple linear regression tasks? We develop new techniques and establish a number of new results in the high-dimensional setting, where the sample size and feature dimension increase at a fixed ratio. First, we show a sharp bias-variance decomposition of hard parameter sharing, given multiple tasks with the same features. Second, we characterize the asymptotic bias-variance limit for two tasks, even when they have arbitrarily different sample size ratios and covariate shifts. We also demonstrate that these limiting estimates for the empirical loss are incredibly accurate in moderate dimensions. Finally, we explain an intriguing phenomenon where increasing one task's sample size helps another task initially by reducing variance but hurts eventually due to increasing bias. This suggests progressively adding data for optimizing hard parameter sharing, and we validate its efficiency in text classification tasks.
\end{abstract}

@@

{\cor rewrite} Hard parameter sharing (HPS) for multi-task learning is widely used in empirical research and goes back to the seminal work of \cite{C97}.
Recent work has revived interests in this approach because it improves performance and reduces the cost of collecting labeled data \cite{R17}.
It is generally applied by sharing the feature layers between all tasks while keeping an output layer for every task.
Often, hard parameter sharing offers two critical advantages if successfully applied.
First, it reduces model parameters since all tasks use the same feature space.
Second, it reduces the amount of labeled data needed from each task by augmenting the entire training dataset.

@@

\section{Transfer learning for high-dimensional linear regression}\label{sec_HPS}
In this section, we define our model for the setting of transfer learning, and introduce several transfer learning estimators that will be considered in this paper.

\subsection{The model}
Consider the transfer learning setting with two data sets. We denote their sample sizes by $n_1$ and $n_2$. For $i=1,2,$ Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates. We assume that the label vector $Y^{(i)} \in \real^{n_i}$ for $X^{(i)}$ follows a linear model plus random noise:
$$Y^{(i)}= X^{(i)}\beta^{(i)} + \varepsilon^{(i)},\quad i=1,2.$$
In this paper, we make the following assumptions on feature covariates $X^{(i)}$ and the noise vectors $\varepsilon^{(i)}$. We remark that all these assumptions are natural in high-dimensional statistics.

@@

\subsection{Transfer learning estimators}

\FY{Introduce some popular transfer learning estimators: total loss estimator, and model averaging estimator.}

We study the standard hard parameter sharing architecture: a shared feature representation layer $B\in\real^{p}$ for all datasets and a separate output layer $A_i \in \real$ for every dataset $i$. Then we study the following minimization problem:
\begin{align}\label{eq_tsl}
			f(A, B) = \norm{X^{(1)} B A_1 - Y^{(1)}}^2 + \norm{X^{(2)} B A_2 - Y^{(2)}}^2,
\end{align}
where we abbreviate $A = [A_1, A_2]$. Let $(\hat{A}, \hat{B})$ be the minimizer of $f(A, B)$. We define the hard parameter sharing (HPS) estimator for task $i$ as 
\be\label{def_HPS}\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i,\quad i=1,2.\ee

@@

{\cor discuss about the bias-variance trade-off}
%However, to the best of our knowledge, the asymptotic limit of $L(\hat{\beta}_2^{\MTL}(a))$ has never been calculated exactly in any high-dimensional setting.

\section{Predication loss for HPS estimator}

\subsection{Varying sample sizes} \label{sec_sizeratio}

@@

For a fixed $n_2$, we study the intricate bias-variance tradeoff with respect to $n_1$ using the following function of $\rho:=\rho_1+\rho_2$,
$$h(\rho):=\frac{p }{n_2-p}\cdot \frac{\sigma^2}{2d^2}- \frac{\ell(\rho_1p , \rho_2p)}{2d^2}  = \frac{\rho-\rho_2}{(\rho-1)(\rho_2-1)}\cdot \frac{\sigma^2}{2d^2}- \frac{(\rho-\rho_2)^2 (\rho-1)+(\rho-\rho_2)\rho_2}{\rho^2(\rho-1)}  .$$
This function characterizes $L(\hat{\beta}_2^{\STL} )-L(\hat{\beta}_2^{\MTL}(\hat a)) $, which gives the quantitive information transfer from the source task to the target task.
First, we have positive (resp. negative) transfer if and only if $h(\rho)>0$ (resp. $h(\rho)<0$). Moreover, the sign of $h(\rho)$ is determined by the sign of the second order polynomial %which is equivalent to the following inequality:
\be\label{poly1} \left( \frac{1}{\rho_2-1}\cdot \frac{\sigma^2}{2d^2}-1 \right)\rho^2 + ( \rho_2+1)\rho - 2\rho_2 .\ee
Then we observe the following dichotomy. \FY{add several plots to illustrate this dichotomy}
\begin{enumerate}
\item If $\frac{1}{\rho_2-1}\cdot \frac{\sigma^2}{2d^2}-1 >0$, then the polynomial \eqref{poly1} is positive for all $\rho\in (\rho_2+1,\infty)$, so that the transfer is always positive.

@@

\item If $(1+\rho_2)^2 < 8\rho_2 \left( 1- \frac{1}{\rho_2-1}\cdot \frac{\sigma^2}{2d^2}  \right)$, the polynomial \eqref{poly1} has two positive roots, where one of them is always smaller than $\rho_2+1$. Hence if the larger root, say $\rho_c$, is smaller than $\rho_2+1$, then the transfer is always negative for $\rho\in (\rho_2+1,\infty)$; otherwise, there is a transition from positive transfer to negative transfer as $\rho$ crosses $\rho_c$.
\end{itemize}
\end{enumerate}
Second, we calculate the derivative of $h(\rho)$, and find that its sign  is determined by the sign of the third order polynomial 
\be\label{poly2}
 \frac{1}{\rho_2}\cdot \frac{\sigma^2}{2d^2} \rho^3 - 2(\rho-\rho_2)(\rho-1)(\rho-2)   - (\rho_2-1)\rho .\ee
Then we observe the following dichotomy. \FY{add several plots to illustrate this dichotomy}
\begin{enumerate}
\item If $\frac{1}{\rho_2}\cdot \frac{\sigma^2}{2d^2} > 2$, the polynomial \eqref{poly2} is always positive for all $\rho\in (\rho_2+1,\infty)$, so that the information transfer always increases as $n_1$ increases.

\item If $\frac{1}{\rho_2}\cdot \frac{\sigma^2}{2d^2} < 2$, the polynomial \eqref{poly2} is positive initially around $\rho=\rho_2+1$, and then becomes negative when $\rho$ crosses its unique root, say $\rho_c$, in $(\rho_2+1,\infty)$. Hence the information transfer achieves the global maximum at $\rho=\rho_c$.
\end{enumerate}

@@

\subsection{Covariate shift}\label{sec_diff}

@@

\section{Extension to multi-task learning}\label{sec_same}

@@

% \section*{Acknowledgements}
%Thanks to Edgar Dobriban, Vaggos Chatziafratis, Mayee Chen, and Ruoxuan Xiong for helpful discussions at various stages of this work. Hongyang R. Zhang and Weijie J. Su's contributions were in part supported by the Wharton Dean's Fund for Postdoctoral Research.
%We gratefully acknowledge the support of DARPA under Nos. FA86501827865 (SDH) and FA86501827882 (ASED); NIH under No. U54EB020405 (Mobilize), NSF under Nos. CCF1763315 (Beyond Sparsity), CCF1563078 (Volume to Velocity), and 1937301 (RTML); ONR under No. N000141712266 (Unifying Weak Supervision); the Moore Foundation, NXP, Xilinx, LETI-CEA, Intel, IBM, Microsoft, NEC, Toshiba, TSMC, ARM, Hitachi, BASF, Accenture, Ericsson, Qualcomm, Analog Devices, the Okawa Foundation, American Family Insurance, Google Cloud, Swiss Re, the HAI-AWS Cloud Credits for Research program, and members of the Stanford DAWN project: Teradata, Facebook, Google, Ant Financial, NEC, VMWare, and Infosys. The U.S. Government is authorized to reproduce and distribute reprints for Governmental purposes notwithstanding any copyright notation thereon. Any opinions, findings, and conclusions or recommendations expressed in this material are those of the authors and do not necessarily reflect the views, policies, or endorsements, either expressed or implied, of DARPA, NIH, ONR, or the U.S. Government.
